Multi-State Temporal Occupancy Modeling

### 2020-07-30 Model Statement/MSTOM Statement.tex
edits based on mason review
alter pdf

Below, we outline a series of multi-state temporal occupancy models (MSTOM) that can be used to understand temporal habitat-use. This model is a special form of the general multi-state occupancy model with state uncertainty \citep{MacKenzie2009}.

@@

\subsubsection{Multinomial-logit Model 1}
To model site-level occurrence according to a single continuous covariates ($\mathbf{x}$) that effects state-level probabilities differently, we could specify this as,
\begin{center}
\begin{align*}
\textbf{z}_{i} &\sim \text{Categorical}(\boldsymbol{\psi_{i}})\\
\boldsymbol{\psi}_{i} &= \begin{bmatrix} \psi^1_{i} & \psi^2_{i} & \psi^3_{i} & \psi^4_{i} \end{bmatrix}.\\
\psi^1_{i} &=\frac{\phi^1_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}\\
\psi^2_{i} &=\frac{\phi^2_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}\\
\psi^3_{i} &=\frac{\phi^3_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}\\
\psi^4_{i} &=\frac{\phi^4_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}.\\
\end{align*}
\end{center}

\begin{center}
\begin{align*}
\phi^1_{i} &= 1\\
\phi^2_{i} &= e^{\alpha_{1}+\alpha_{2}\times x_{i}}\\
\phi^3_{i} &= e^{\alpha_{3}+\alpha_{4}\times x_{i}}\\
\phi^4_{i} &= e^{\alpha_{5}+\alpha_{6}\times x_{i}}\\
\end{align*}
\end{center}

@@

\subsubsection{Multinomial-logit Model 2}
We could generalize the modeling of state-specific occupancy parameters using separate design matrices ($\textbf{x}_{i}^{\text{Day}}, \textbf{x}_{i}^{\text{Night}}, \textbf{x}_{i}^{\text{ND}}$) that for each site $i$ are  1 x Q$_{m}$ (being the number of columns) and associated vectors of coefficients ($\boldsymbol{\alpha}^{\text{Day}}, \boldsymbol{\alpha}^{\text{Night}}, \boldsymbol{\alpha}^{\text{ND}}$) that are $Q_{m}$ x 1, and specified as,
\begin{center}
\begin{align*}
\textbf{z}_{i} &\sim \text{Categorical}(\boldsymbol{\psi_{i}})\\
\boldsymbol{\psi}_{i} &= \begin{bmatrix} \psi^1_{i} & \psi^2_{i} & \psi^3_{i} & \psi^4_{i} \end{bmatrix}.\\
\psi^1_{i} &=\frac{\phi^1_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}\\
\psi^2_{i} &=\frac{\phi^2_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}\\
\psi^3_{i} &=\frac{\phi^3_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}\\
\psi^4_{i} &=\frac{\phi^4_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}.\\
\end{align*}
\end{center}

\begin{center}
\begin{align*}
\phi^1_{i} &= 1\\
\phi^2_{i} &= e^{\textbf{x}_{i}^{\text{Day}}\boldsymbol{\alpha}^{\text{Day}}}\\
\phi^3_{i} &= e^{\textbf{x}_{i}^{\text{Night}}\boldsymbol{\alpha}^{\text{Night}}}\\
\phi^4_{i} &= e^{\textbf{x}_{i}^{\text{ND}}\boldsymbol{\alpha}^{\text{ND}}}\\
\end{align*}
\end{center}

@@

This model could be easily extended to include a difference of occupancy and detection in state 4 by a small modification of $\phi^4_{i} = e^{\alpha_{1}+\alpha_{2}+\alpha_{3}}$ and the matrix cell $q^{4,4}$ to $e^{\beta_{1}+\beta_{2}+\beta_{3}} $. A diffuse prior for all coefficients could be done using $\text{Logistic}(0,1).$ This model is coded for JAGS in the file `jags.multistate.occ.reduced.alt.R'.\\

@@

>>>>

Below, we outline a series of multi-state temporal occupancy models (MSTOM) that can be used to understand temporal habitat-use. This model is a special form of the general multi-state occupancy model with state uncertainty \citep{MacKenzie2009}. Furthermore, the MSTOM Full model is equivalent to the original co-occurence occupancy model \citep{MacKenzie2004} with two-species and the equivalent four states. A demonstration of their equivalent estimation of the most probable parameters values (i.e., maximum likelihood estimates or mode of the posterior distribution) can be found in the R script, `MSTOM Full comparison with 2.species.occ.model.r'. Note, that while MacKenzie et al. 2004 is known to have numerical optimization issues, especially when modeling covariates \citep{MacKenzie2004}, this issue does not persist when using a hierarchical Bayesian approach with markov chain monte carlo optimization.

@@

\subsubsection{Multinomial-logit Model 1}
Instead of focusing on estimating site-level occupancy probabilities, we could recast this model using a multinomial logit link as, 
\begin{center}
\begin{align*}
\textbf{z}_{i} &\sim \text{Categorical}(\boldsymbol{\psi})\\
\boldsymbol{\psi}_{i} &= \begin{bmatrix} \psi^1_{i} & \psi^2_{i} & \psi^3_{i} & \psi^4_{i} \end{bmatrix}.\\
\psi^1 &=\frac{\phi^1_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}\\
\psi^2 &=\frac{\phi^2_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}\\
\psi^3 &=\frac{\phi^3_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}\\
\psi^4 &=\frac{\phi^4_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}.\\
\end{align*}
\end{center}

\begin{center}
\begin{align*}
\phi^1 &= 1\\
\phi^2 &= e^{\alpha_{1}}\\
\phi^3 &= e^{\alpha_{2}}\\
\phi^4 &= e^{\alpha_{1}+\alpha_{2}+\alpha_{3}}\\
\end{align*}
\end{center}

A diffuse prior for all coefficients could be done using $\text{Logistic}(0,1).$ This model is coded for JAGS in the file `jags.multistate.occ.full.alt.R'.

\subsubsection{Multinomial-logit Model 2}
To model site-level occurrence according to a single continuous covariates ($\mathbf{x}$) that effects state-level probabilities differently, we could specify this as,
\begin{center}
\begin{align*}
\textbf{z}_{i} &\sim \text{Categorical}(\boldsymbol{\psi_{i}})\\
\boldsymbol{\psi}_{i} &= \begin{bmatrix} \psi^1_{i} & \psi^2_{i} & \psi^3_{i} & \psi^4_{i} \end{bmatrix}.\\
\psi^1_{i} &=\frac{\phi^1_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}\\
\psi^2_{i} &=\frac{\phi^2_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}\\
\psi^3_{i} &=\frac{\phi^3_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}\\
\psi^4_{i} &=\frac{\phi^4_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}.\\
\end{align*}
\end{center}

\begin{center}
\begin{align*}
\phi^1_{i} &= 1\\
\phi^2_{i} &= e^{\alpha_{1}+\alpha_{2}\times x_{i}}\\
\phi^3_{i} &= e^{\alpha_{3}+\alpha_{4}\times x_{i}}\\
\phi^4_{i} &= e^{\alpha_{1}+\alpha_{2}\times x_{i}+\alpha_{3}+\alpha_{4}\times x_{i}+\alpha_{5}+\alpha_{6}\times x_{i}}\\
\end{align*}
\end{center}

@@

\subsubsection{Multinomial-logit Model 3}
We could generalize the modeling of state-specific occupancy parameters using separate design matrices ($\textbf{x}_{i}^{\text{Day}}, \textbf{x}_{i}^{\text{Night}}, \textbf{x}_{i}^{\text{ND}}$) that for each site $i$ are  1 x Q$_{m}$ (being the number of columns) and associated vectors of coefficients ($\boldsymbol{\alpha}^{\text{Day}}, \boldsymbol{\alpha}^{\text{Night}}, \boldsymbol{\alpha}^{\text{ND}}$) that are $Q_{m}$ x 1, and specified as,
\begin{center}
\begin{align*}
\textbf{z}_{i} &\sim \text{Categorical}(\boldsymbol{\psi_{i}})\\
\boldsymbol{\psi}_{i} &= \begin{bmatrix} \psi^1_{i} & \psi^2_{i} & \psi^3_{i} & \psi^4_{i} \end{bmatrix}.\\
\psi^1_{i} &=\frac{\phi^1_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}\\
\psi^2_{i} &=\frac{\phi^2_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}\\
\psi^3_{i} &=\frac{\phi^3_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}\\
\psi^4_{i} &=\frac{\phi^4_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}.\\
\end{align*}
\end{center}

\begin{center}
\begin{align*}
\phi^1_{i} &= 1\\
\phi^2_{i} &= e^{\textbf{x}_{i}^{\text{Day}}\boldsymbol{\alpha}^{\text{Day}}}\\
\phi^3_{i} &= e^{\textbf{x}_{i}^{\text{Night}}\boldsymbol{\alpha}^{\text{Night}}}\\
\phi^4_{i} &= e^{\textbf{x}_{i}^{\text{Day}}\boldsymbol{\alpha}^{\text{Day}}+\textbf{x}_{i}^{\text{Night}}\boldsymbol{\alpha}^{\text{Night}}+\textbf{x}_{i}^{\text{ND}}\boldsymbol{\alpha}^{\text{ND}}}\\
\end{align*}
\end{center}

@@

This model could be easily extended to include a difference of occupancy and detection in state 4 by a small modification of $\phi^4_{i} = e^{\alpha_{1}+\alpha_{2}+\alpha_{3}}$ and the matrix cell $q^{4,4}$ to $e^{\beta_{1}+\beta_{2}+\beta_{3}} $. A diffuse prior for all coefficients could be done using $\text{Logistic}(0,1).$ \\

@@

\bibitem[Mackenzie et al., 2004]{MacKenzie2004}MacKenzie, D. I., Bailey, L. L., \& Nichols, J. D. (2004). Investigating species co‐occurrence patterns when species are detected imperfectly. Journal of Animal Ecology, 73(3), 546-555.
